\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-14 NAR_DRAFT.tex
morning

\batcave is a post-call variant evalutation tool that 
We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

We have chosen to implement \batcave as a post-calling algorithm for use with MuTect because it presents a simple model and a well-tested set of heuristic filters.
However, in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
In general we are advancing the idea that there is substantial prior information about the tumor and site-specific mutation probability encoded in NGS data, and we should be using it.

>>>>

\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

The \batcave algorithm focuses on the tri-nucleotide context and substitution type, but there are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
These other types of context could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
However, in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
In general we are advancing the idea that there is substantial prior information about the tumor and site-specific mutation probability encoded in NGS data, and we should be using it.
